Annotation-Based Empirical Performance Tuning Using Orio

### 2009-02-12 doc/papers/ipdps09/integration.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@428 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@433 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

@@

Figure~\ref{fig:pluto-orio} outlines the overall structure of the Pluto-Orio
integrated system, which is implemented as a new optimization module in
Orio.
%% BN: the detailed description below is good, but commented out for now for space.
%Initially, the transformation process takes two kinds of input:
%the loop code to be optimized and the optimization parameters code. The loop
%code is embedded in the annotation body block, while the performance
%parameters code is written in the module body block. A simple parser
%implemented inside the new module parses the performance parameters code to
%extract their values. 
In addition to the loop code to be optimized, the module parses performance
parameters, which include tile sizes, unroll factors, loop order
(permutation), as well as several boolean values for triggering OpenMP
parallelization, scalar replacement, and auto-vectorization. Using these
parameter values, Pluto then performs polyhedral transformations (e.g.,
two-level tiling and OpenMP parallelization) on the input loop code. The
resulting generated code is subsequently passed to the widely-available
profiling tool gprof~\cite{gprof} for hotspot detection. 
%The statistical
%information produced by gprof provides accurate locations (line numbers) of
%the loop nests where the Pluto-generated code spends most of its execution
%time. 
The identified hotspot loop nests are then decorated with Orio's performance
annotations for complementary syntactic transformations (e.g., loop
permutation, loop unroll/jamming, scalar replacement, and explicit
auto-vectorization). Finally, Orio optimizes the annotated hotspots as 
described in Section~\ref{sec:implm}.

\begin{figure}[tb]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/pluto-orio.eps}    
\end{center}   
\vspace{-.2in}
\caption{Integration of Pluto into Orio's code transformation module.}
\label{fig:pluto-orio}
\end{figure}

>>>>

\begin{figure*}[bt]
\begin{center}
\includegraphics[width=0.6\textwidth]{figures/pluto-orio.eps}    
\end{center}   
\caption{Integration of Pluto into Orio's code transformation module.}
\label{fig:pluto-orio}
\end{figure*}

@@

Figure~\ref{fig:pluto-orio} outlines the overall structure of the
Pluto-Orio integrated system, which is implemented as a new
optimization module in Orio. Initially, the transformation process
takes two kinds of input: the loop code to be optimized and the
optimization parameters code. The loop code is embedded in the
annotation body block, while the performance parameters code is
written in the module body block. A simple parser implemented inside
the new module parses the performance parameters code to extract their
values. In addition to the loop code to be optimized, the module
parses performance parameters, which include tile sizes, unroll
factors, loop order (permutation), as well as several boolean values
for triggering OpenMP parallelization, scalar replacement, and
auto-vectorization. Using these parameter values, Pluto then performs
polyhedral transformations (e.g., two-level tiling and OpenMP
parallelization) on the input loop code. The resulting generated code
is subsequently passed to the widely-available profiling tool
gprof~\cite{gprof} for hotspot detection. The statistical information
produced by gprof provides accurate locations (line numbers) of the
loop nests where the Pluto-generated code spends most of its execution
time. The identified hotspot loop nests are then decorated with Orio's
performance annotations for complementary syntactic transformations
(e.g., loop permutation, loop unroll/jamming, scalar replacement, and
explicit auto-vectorization). Finally, Orio optimizes the annotated
hotspots as described in Section~\ref{sec:implm}.

### 2009-02-12 doc/papers/ipdps09/related.tex
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@428 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb
git-svn-id: https://svn.mcs.anl.gov/repos/performance/orio@433 e7ad4b5f-b827-0410-872f-f7f4bc3d1efb

Ideally, a developer should only have to specify a few simple command-line
options and then rely on the compiler to optimize the performance of an
application on any architecture. Compilers alone, however, cannot fully
satisfy the performance needs of scientific applications.  First,
compilers must operate in a black-box fashion and at a very low level,
limiting both the type and number of optimizations that can be done.  Second,
static analysis of general-purpose languages, such as C, C++, and Fortran, is
necessarily conservative, thereby precluding many possible optimizations.
%Third, in the process of transforming a mathematical model into a computer
%program, much potentially useful (for optimization purposes) information is
%lost since it cannot be represented by the programming language.  
Finally, extensive manual tuning of a code may prevent certain compiler optimizations
and result in worse performance on new architectures, resulting in 
loss of performance portability.

%As briefly discussed in Section~\ref{sec:motivation}, performance tuning is
%generally approached in three ways: by performing manual optimizations of key
%portions of the code; by using compiler-based source transformation tools for
%loop optimizations; and by using tuned libraries for key numerical
%algorithms.
%% Libraries
An alternative to manual or automated tuning of application codes is the use
of tuned libraries. The two basic approaches to supplying high-performance
libraries include providing a library of hand-coded options (e.g.,
\cite{BLAS,ESSL,Goto:2006fk}) and generating optimized code automatically
for the given problem and machine parameters. 
ATLAS~\cite{atlas_sc98}
%,WN147} 
for BLAS and LAPACK
%~\cite{BLAS} and LAPACK~\cite{laug}
routines, OSKI~\cite{OSKI} for sparse linear algebra,
PHiPAC~\cite{bilmes97optimizing} for matrix-matrix products, and
domain-specific libraries such as FFTW~\cite{frigo98} and
SPIRAL~\cite{SPIRAL} are all examples of the latter approach. Most 
automatic tuning approaches perform empirical
parameter searches on the target platform.  
%FFTW uses a combination of static
%models and empirical techniques to optimize FFTs. SPIRAL generates optimized
%digital signal processing libraries by an extensive empirical search over
%implementation variants.  GotoBLAS~\cite{Goto:2006fk,Goto:fk}, on the other
%hand, achieves great performance results on several architectures by using
%hand-tuned data structures and kernel operations.  
These auto- or hand-tuned
approaches can deliver performance that can be five times as fast as that
produced by many optimizing compilers \cite{WN147}.  The library approach,
however, is limited by the fact that optimizations are highly problem- and
machine-dependent. Furthermore, at this time, the functionality of the
currently available automated tuning systems is quite limited.

%% Other annotation-based source transformation approaches
General-purpose tools for optimizing loop performance are also available.
LoopTool~\cite{LoopTool} supports annotation-based loop fusion,
unroll/jamming, skewing and tiling.  The Matrix Template Library
\cite{Siek:1998ys} uses template metaprograms to tile at both the register
and cache levels.  A new tool, POET~\cite{POET} also supports a number of
loop transformations. 
%POET offers a complex template-based syntax for
%defining transformations in a language-independent manner. 
Other research efforts whose goal, at least in part, is to enable
optimizations of source code to be augmented with performance-related
information include the X language~\cite{XLanguage} (a macro C-like language
for annotating C code), the Broadway~\cite{broadway} compiler, and
telescoping languages~\cite{telescopingurl}.
%,teleoverview,Ken99}
%, and various
%meta-programming techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}.

\comment{
Emerging annotation-based tools are normally designed by compiler researchers
and thus the interfaces are not necessarily based on concepts accessible to
computational scientists. The complexity of existing annotation languages and lack
of common syntaxes for transformations (e.g., loop unrolling) result
in steep learning curves and the inability to take advantage of more than one
approach at a time. Furthermore, at present, there is no good way for
\emph{users} to learn about the tools available and compare their
capabilities and performance.
}

>>>>

Ideally, a developer should only have to specify a few simple
command-line options and then rely on the compiler to optimize the
performance of an application on any architecture. Compilers alone,
however, cannot fully satisfy the performance needs of scientific
applications.  First, compilers must operate in a black-box fashion
and at a very low level, limiting both the type and number of
optimizations that can be done.  Second, static analysis of
general-purpose languages, such as C, C++, and Fortran, is necessarily
conservative, thereby precluding many possible optimizations.  Third,
in the process of transforming a mathematical model into a computer
program, much potentially useful (for optimization purposes)
information is lost since it cannot be represented by the programming
language.  Finally, extensive manual tuning of a code may prevent
certain compiler optimizations and result in worse performance on new
architectures, resulting in loss of performance portability.

%As briefly discussed in Section~\ref{sec:motivation}, performance tuning is
%generally approached in three ways: by performing manual optimizations of key
%portions of the code; by using compiler-based source transformation tools for
%loop optimizations; and by using tuned libraries for key numerical
%algorithms.
%% Libraries

An alternative to manual or automated tuning of application codes is the use
of tuned libraries. The two basic approaches to supplying high-performance
libraries include providing a library of hand-coded options (e.g.,
\cite{BLAS,ESSL,Goto:2006fk}) and generating optimized code automatically
for the given problem and machine
parameters. ATLAS~\cite{atlas_sc98,WN147} for BLAS~\cite{BLAS} and
some LAPACK~\cite{laug} routines, OSKI~\cite{OSKI} for sparse linear
algebra, PHiPAC~\cite{bilmes97optimizing} for matrix-matrix products,
and domain-specific libraries such as FFTW~\cite{frigo98} and
SPIRAL~\cite{SPIRAL} are all examples of the latter approach. Most
automatic tuning approaches perform empirical parameter searches on
the target platform.  FFTW uses a combination of static models and
empirical techniques to optimize FFTs. SPIRAL generates optimized
digital signal processing libraries by an extensive empirical search
over implementation variants.  GotoBLAS~\cite{Goto:2006fk,Goto:fk}, on
the other hand, achieves great performance results on several
architectures by using hand-tuned data structures and kernel
operations.  These auto- or hand-tuned approaches can deliver
performance that can be five times as fast as that produced by many
optimizing compilers
\cite{WN147}.  The library approach, however, is limited by the fact
that optimizations are highly problem- and
machine-dependent. Furthermore, at this time, the functionality of the
currently available automated tuning systems is quite limited.

%% Other annotation-based source transformation approaches

General-purpose tools for optimizing loop performance are also
available.  LoopTool~\cite{LoopTool} supports annotation-based loop
fusion, unroll/jamming, skewing and tiling.  The Matrix Template
Library \cite{Siek:1998ys} uses template metaprograms to tile at both
the register and cache levels.  A new tool, POET~\cite{POET} also
supports a number of loop transformations. POET offers a complex
template-based syntax for defining transformations in a
language-independent manner. Other research efforts whose goal, at
least in part, is to enable optimizations of source code to be
augmented with performance-related information include the X
language~\cite{XLanguage} (a macro C-like language for annotating C
code), the Broadway~\cite{broadway} compiler, and telescoping
languages~\cite{telescopingurl,teleoverview,Ken99}, and various
meta-programming
techniques~\cite{veldhuizen95,weise93,kiczales91,chiba95}.

Emerging annotation-based tools are normally designed by compiler researchers
and thus the interfaces are not necessarily based on concepts accessible to
computational scientists. The complexity of existing annotation languages and lack
of common syntaxes for transformations (e.g., loop unrolling) result
in steep learning curves and the inability to take advantage of more than one
approach at a time. Furthermore, at present, there is no good way for
\emph{users} to learn about the tools available and compare their
capabilities and performance.
